Weakly Nonlinear Analysis of the Standard and Helical Magnetorotational Instabilities in a Taylor Couette Flow

### 2016-08-01 paper/widegap_mri_paper.tex
more widegap paper words

\begin{align}
& \partial_t \uphi + \frac{1}{r^2} u_0 \partial_z \Psi + \frac{1}{r} \partial_r u_0 \partial_z \Psi - \frac{2}{\beta} B_0 \partial_z B_\phi - \frac{1}{\reye} ( \nabla^2 \uphi - \frac{1}{r^2} \uphi ) = N^{u} \label{eq:uphi_perturbed} \\
& \partial_t A - B_0 \partial_z \Psi - \frac{1}{\reym} ( \nabla^2 A - \frac{2}{r} \partial_r A )= N^{(A)}  \label{eq:A_perturbed}\\
  \label{eq:Bphi_perturbed}
& \partial_t B_\phi + \frac{1}{r^2} u_0 \partial_z A - B_0 \partial_z u_\phi - \frac{1}{r} \partial_r u_0 \partial_z A - \frac{2}{r^3} B_0 \xi \partial_z \Psi - \frac{1}{\reym} (\nabla^2 B_\phi - \frac{1}{r^2} B_\phi ) = N^{(B)}
\end{align}

@@

\beq
N^{(\Psi)} = - J(\Psi, \frac{1}{r^2} ( \nabla^2 \Psi - \frac{2}{r} \partial_r\Psi) ) + \frac{2}{\beta} J(A, \frac{1}{r^2} ( \nabla^2 A - \frac{2}{r} \partial_rA) ) - \frac{2}{\beta} \frac{2}{r}B_\phi \partial_z B_\phi  + \frac{2}{r} u_\phi \partial_z u_\phi 
\eeq
\beq
N^{(u)} = \frac{2}{\beta} \frac{1}{r} J(A, B_\phi) - \frac{1}{r} J(\Psi, \uphi) + \frac{2}{\beta}\frac{1}{r^2} B_\phi \partial_z A - \frac{1}{r^2} \uphi \partial_z \Psi 
\eeq
\beq
N^{(A)} = \frac{1}{r} J(A, \psi)
\eeq
\beq
N^{(B)} = \frac{1}{r} J(A, \uphi) + \frac{1}{r} J(B_\phi, \psi) + \frac{1}{r^2} B_\phi \partial_z \psi - \frac{1}{r^2} \uphi \partial_z A 
\eeq

@@

where J is the Jacobian $J(f, g) \equiv \partial_z f \partial_r g - \partial_r f \partial_z g$. Note that in the above, $\nabla^2 f \equiv \partial_r^2 f + \partial_z^2 f + \frac{1}{r} \partial_r f$. Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} are nondimensionalized, where lengths have been scaled by $r_0$, velocities by $r_0 \Omega_0$, densities by $\rho_0$, and magnetic fields by $B_0$; where $B_0$ appears in the above it is formally unity. $\Omega_0 = \Omega(r_0)$ is the rotation rate at the center of the channel. We introduce the Reynolds number $\reye = \Omega_0 r_0^2/\nu$, the magnetic Reynolds number $\reym = \Omega_0 r_0^2 / \eta$, and a plasma beta parameter $\beta = \Omega_0^2 r_0^2 \rho_0/B_0^2$. If we define the dimensional cylindrical coordinate $r = r_0(1 + \delta x)$, the narrow gap equations are recovered in the limit $\delta \rightarrow 0$.

We solve the system subject to periodic vertical boundary conditions and no-slip, perfectly conducting radial boundary conditions.

@@

See Appendix \ref{app:basic_equations} for the definition of matrices and nonlinear vectors, and a thorough derivation. We emphasize that Equations \ref{eq:ordere} - \ref{eq:ordere3} have the same form as these equations in the narrow gap case, although the matrices (which contain all radial derivatives) are significantly different in this wide gap formulation. This is because we do not have slow variation in the radial dimension. The slow variation in $Z$ and $T$ are parameterized as an amplitude function $\alpha(Z, T)$ which modulates the flow in these dimensions. This parameterization coupled with the boundary conditions lead us to an ansatz linear solution $\mathbf{V}_1 = \alpha(Z, T) \mathbb{V}_{11}(r) e^{i k_z z} + c.c.$, where the radial variation is contained in $\mathbb{V}_{11}$. We defer full expressions and derivations thereof to the Appendix, and focus here on features of the wide gap system which distinguish it from the narrow gap limit.

>>>>

\begin{align}
& \partial_t \uphi + \frac{1}{r^2} u_0 \partial_z \Psi + \frac{1}{r} \partial_r u_0 \partial_z \Psi - \frac{2}{\beta} B_0 \partial_z B_\phi - \frac{1}{\reye} ( \nabla^2 \uphi - \frac{1}{r^2} \uphi ) = N^{(u)} \label{eq:uphi_perturbed} \\
& \partial_t A - B_0 \partial_z \Psi - \frac{1}{\reym} ( \nabla^2 A - \frac{2}{r} \partial_r A )= N^{(A)}  \label{eq:A_perturbed}\\
  \label{eq:Bphi_perturbed}
& \partial_t B_\phi + \frac{1}{r^2} u_0 \partial_z A - B_0 \partial_z u_\phi - \frac{1}{r} \partial_r u_0 \partial_z A - \frac{2}{r^3} B_0 \xi \partial_z \Psi - \frac{1}{\reym} (\nabla^2 B_\phi - \frac{1}{r^2} B_\phi ) = N^{(B)}
\end{align}

@@

\beq
N^{(\Psi)} = - J(\Psi, \frac{1}{r^2} ( \nabla^2 \Psi - \frac{2}{r} \partial_r\Psi) ) + \frac{2}{\beta} J(A, \frac{1}{r^2} ( \nabla^2 A - \frac{2}{r} \partial_rA) ) - \frac{2}{\beta} \frac{2}{r}B_\phi \partial_z B_\phi  + \frac{2}{r} u_\phi \partial_z u_\phi 
\eeq
\beq
N^{(u)} = \frac{2}{\beta} \frac{1}{r} J(A, B_\phi) - \frac{1}{r} J(\Psi, \uphi) + \frac{2}{\beta}\frac{1}{r^2} B_\phi \partial_z A - \frac{1}{r^2} \uphi \partial_z \Psi 
\eeq
\beq
N^{(A)} = \frac{1}{r} J(A, \psi)
\eeq
\beq
N^{(B)} = \frac{1}{r} J(A, \uphi) + \frac{1}{r} J(B_\phi, \psi) + \frac{1}{r^2} B_\phi \partial_z \psi - \frac{1}{r^2} \uphi \partial_z A 
\label{eq:nonlinear_B}
\eeq

@@

where J is the Jacobian $J(f, g) \equiv \partial_z f \partial_r g - \partial_r f \partial_z g$. Note that in the above, $\nabla^2 f \equiv \partial_r^2 f + \partial_z^2 f + \frac{1}{r} \partial_r f$. Equations \ref{eq:Psi_perturbed} - \ref{eq:nonlinear_B} are nondimensionalized, where lengths have been scaled by $r_0$, velocities by $r_0 \Omega_0$, densities by $\rho_0$, and magnetic fields by $B_0$; where $B_0$ appears in the above it is formally unity. $\Omega_0 = \Omega(r_0)$ is the rotation rate at the center of the channel. We introduce the Reynolds number $\reye = \Omega_0 r_0^2/\nu$, the magnetic Reynolds number $\reym = \Omega_0 r_0^2 / \eta$, and a plasma beta parameter $\beta = \Omega_0^2 r_0^2 \rho_0/B_0^2$. If we define the dimensional cylindrical coordinate $r = r_0(1 + \delta x)$, the narrow gap equations are recovered in the limit $\delta \rightarrow 0$.

We solve the system subject to periodic vertical boundary conditions and no-slip, perfectly conducting radial boundary conditions, namely

\beq
\Psi = \partial_r \Psi = u = A = \partial_r (r B) = 0
\eeq

at $r = r_1, r_2$.

We note that Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} are written in a nonstandard form, with the nonlinear terms on the righthand side. This choice has a practical motivation. As detailed in \S\ref{sec:wnl_analysis}, we expand these equations in a perturbation series and solve them order by order using a pseudospectral code. The code solves partial differential equations of the form $\mathrm{M} \partial_t \mathbf{V} + \mathrm{L} \mathbf{V} = \mathbf{F}$, where $\mathrm{M}$ and $\mathrm{L}$ are matrices and $\mathbf{F}$ is a vector containing any inhomogenous terms. The nonlinear terms in our perturbation analysis become inhomogenous term inputs to the solver.

@@

See Appendix \ref{app:basic_equations} for the definition of matrices and nonlinear vectors, and a thorough derivation. We emphasize that Equations \ref{eq:ordere} - \ref{eq:ordere3} have the same form as these equations in the narrow gap case, although the matrices (which contain all radial derivatives) are significantly different in this wide gap formulation. This is because we do not have slow variation in the radial dimension. The slow variation in $Z$ and $T$ are parameterized as an amplitude function $\alpha(Z, T)$ which modulates the flow in these dimensions. This parameterization coupled with the boundary conditions lead us to an ansatz linear solution $\mathbf{V}_1 = \alpha(Z, T) \mathbb{V}_{11}(r) e^{i k_z z} + c.c.$, where the radial variation is contained in $\mathbb{V}_{11}$.

We solve the equations at each order using Dedalus, an open source pseudospectral code (Burns et al. in prep). Because our domain is bounded only radially and not vertically, we solve the radial dimension on a basis of Chebyshev polynomials, and the vertical dimension on a Fourier series basis. We solve Equation \ref{eq:ordere} as a linear eigenvalue problem, and Equations \ref{eq:ordere2} and \ref{eq:ordere3} as linear boundary value problems. %should i mention spurious eigenvals?

%We defer full expressions and derivations thereof to the Appendix, and focus here on features of the wide gap system which distinguish it from the narrow gap limit.
